\section{On the Equivalence of \LCC and \CP}
\label{sec:equivalence}

In \Cref{sec:introduction} we have anticipated that \LCC is a ``linear version''
of \CP in which sessions are encoded instead of being featured natively. In this
section we substantiate this claim by sketching a straightforward isomorphism
between (well-typed) \CP processes and (well-typed) \LCC processes, up to bound
names. Such tight correspondence can be traced back to the papers by
Kobayashi~\cite{Kobayashi02b} and Dardha et al.~\cite{DardhaGiachinoSangiorgi17}
showing how (binary) sessions can be encoded as chains of \emph{linear}
channels.

In the rest of this section, we detail this correspondence focusing on the
multiplicative/additive connectives. The treatment of quantifiers is similar,
and the other parts of \LCC coincide with those of \CP.

If we let $\ClassicalP$ and $\ClassicalQ$ (in sans-serif font) range over \CP
processes and write $\wtcp\ClassicalP\Context$ for a \CP typing judgment, in \CP
the process forms representing input/output operations on sessions are
$\Join\x\y{}.\ClassicalP$ and $\Case\x{}\ClassicalP\ClassicalQ$ (for inputs) and
$\Fork\x\y{}\ClassicalP\ClassicalQ$ and $\Select\x{\InTag_i}{}.\ClassicalP$ (for
outputs). The key difference between these forms and the corresponding ones in
\LCC is the \emph{absence} of an explicit continuation channel. Instead, the
session endpoint $x$ is being used by the continuation processes $\ClassicalQ$
and $\ClassicalP$ to implement the rest of the session. This is made clear by
the \CP typing rules for these processes, which we recall below~\cite{Wadler14}:
\begin{mathpar}
    \inferrule[\StandardForkRule]{
        \wtcp\ClassicalP{y : A, \ContextC}
        \\
        \wtcp\ClassicalQ{x : B, \ContextD}
    }{
        \wtcp{\Fork\x\y{}\ClassicalP\ClassicalQ}{x : A \Ten B, \ContextC, \ContextD}
    }
    \and
    \inferrule[\StandardSelectRule]{
        \wtcp\ClassicalP{x : A_i, \ContextC}
    }{
        \wtcp{\Select\x{\InTag_i}{}.\ClassicalP}{x : A_1 \Plus A_2, \ContextC}
    }
    ~i\in\set{1,2}
    \and
    \inferrule[\StandardJoinRule]{
        \wtcp\ClassicalP{y : A, x : B, \Context}
    }{
        \wtcp{\Join\x\y{}.\ClassicalP}{x : A \Par B, \Context}
    }
    \and
    \inferrule[\StandardCaseRule]{
        \wtcp\ClassicalP{x : A, \Context}
        \\
        \wtcp\ClassicalQ{x : B, \Context}
    }{
        \wtcp{\Case\x{}\ClassicalP\ClassicalQ}{x : A \With B, \Context}
    }
\end{mathpar}

\newcommand\encode[1]{\varepsilon(#1)}
\newcommand\decode[1]{\delta(#1)}

Every (well-typed) \CP process can be \emph{encoded} into a (well-typed) \LCC
process by introducing an explicit continuation channel that does not clash with
other channel names. In order to express such encoding easily as a function
$\encode\cdot$, following Sano et al.~\cite{SanoKavanaghPientka23} we pick as
the name of the continuation channel the very same name of the session that it
continues. The encoding function is defined by the following equations
\begin{align*}
    \encode{\Fork\x\y{}\ClassicalP\ClassicalQ} & = \Fork\x\y\x{\encode\ClassicalP}{\encode\ClassicalQ} \\
    \encode{\Join\x\y{}.\ClassicalP} & = \Join\x\y\x.\encode\ClassicalP \\
    \encode{\Select\x{\InTag_i}{}.\ClassicalP} & = \Select\x{\InTag_i}\x.\encode\ClassicalP \\
    \encode{\Case\x{}\ClassicalP\ClassicalQ} & = \Case\x\x{\encode\ClassicalP}{\encode\ClassicalQ}
\end{align*}
and extended homomorphically to the other forms of \CP.

Every (well-typed) \LCC process can be \emph{decoded} into a (well-typed) \CP
process by renaming each continuation channel so that it matches the name of the
session it continues. The decoding function $\decode\cdot$ is defined by the
following equations
\begin{align*}
    \decode{\Fork\x\y\z{P}{Q}} & = \Fork\x\y{}{\decode{P}}{\decode{Q\subst\x\z}} \\
    \decode{\Join\x\y\z.P} & = \Join\x\y{}.\decode{P\subst\x\z} \\
    \decode{\Select\x{\InTag_i}\z.P} & = \Select\x{\InTag_i}{}.\decode{P\subst\x\z} \\
    \decode{\Case\x\z{P}{Q}} & = \Case\x{}{\decode{P\subst\x\z}}{\decode{Q\subst\x\z}}
\end{align*}
and extended homomorphically to the other forms of \LCC.

It is now easy to establish that $\encode\cdot$ and $\decode\cdot$ are one the
inverse of the other (up to alpha conversion) and that they preserve typing, in
the sense that:
\begin{itemize}
\item If $\wtcp\ClassicalP\Context$, then $\wtp{\encode\ClassicalP}\Context$;
\item If $\wtp{P}\Context$, then $\wtcp{\decode{P}}\Context$.
\end{itemize}

Note that an \LCC process $P$ and its decoding $\decode{P}$ (or equivalently a
\CP process $\ClassicalP$ and its encoding $\encode\ClassicalP$) are typed
\emph{in the same typing context} (up to the name of channels, for typing
judgments other than the concluding one).
